---
title: "One year of online retail: a small, reproducible analysis in R"
author: "See README.md for how to rebuild this document"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
# All logic lives in R/functions.R, shared with run_analysis.R and the tests -
# this document narrates results; it does not define its own pipeline.
source(file.path("R", "functions.R"))

# SVG figures: crisp at any zoom, text-diffable in review, and they render
# natively in GitHub-flavoured markdown.
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8.5, fig.height = 4.2, dev = "svglite",
  dev.args = list(bg = PAL$surface)
)

cleaned <- cleaned_lines()
lines <- cleaned$lines
con <- build_sqlite(lines)
sql <- run_queries(con)
DBI::dbDisconnect(con)

span <- range(as_date(lines$invoice_ts))
total_revenue <- sum(lines$revenue)
```

## What this is

A deliberately small end-to-end analysis of the [UCI Online
Retail](https://doi.org/10.24432/C5BW33) dataset — every transaction of a
UK-based online giftware wholesaler/retailer from
`r format(span[1], "%d %B %Y")` to `r format(span[2], "%d %B %Y")`. The point
is the working style, not the dataset: documented cleaning rules with a
reconciling audit trail, business questions answered in SQL, time series and
customer analysis in R, and a report that rebuilds from raw data with one
command (`Rscript run_analysis.R`).

## Cleaning: every dropped row is accounted for

The raw file is invoice *lines*, and it mixes real product sales with
cancellations, postage charges, fees, and manual stock corrections. Five rules
are applied in a fixed order; each row is counted against the first rule that
removes it, so the audit reconciles exactly against the input.

The netting rule earns its place: the two largest "orders" of the year
(80,995 and 74,215 units) were both **cancelled within minutes** of being
keyed in. Dropping only the credit notes — the obvious first cut — would
leave those phantom sales in every top-product and top-customer list below;
matching each credit note back to its sale (same customer, product, quantity
and price) removes both sides of the pair. Credits are processed in time order;
each consumes the latest unused sale at or before its timestamp. Same-minute
matches are allowed because the source timestamp has minute precision, with
input row order breaking ties. An earlier credit cannot cancel a future sale.

**Correction (2026-09-05):** enforcing this chronological rule restores 263
sales lines and £6,299.05 compared with the previous report. The old code
matched credit counts over the entire observation window; all tables and
figures below have been regenerated with chronological matching.

```{r cleaning-audit}
knitr::kable(
  cleaned$audit |>
    mutate(share_of_input = percent(share_of_input, accuracy = 0.1)),
  col.names = c("Rule (applied in order)", "Rows dropped", "Share of input"),
  format.args = list(big.mark = ",")
)
```

`r format(cleaned$rows_in, big.mark = ",")` raw rows become
`r format(cleaned$rows_out, big.mark = ",")` clean sales lines
(`r percent(1 - cleaned$rows_out / cleaned$rows_in, accuracy = 0.1)` dropped),
worth `r label_gbp(total_revenue)` in revenue. Two caveats worth stating up
front:

- **Netting is exact-match only.** A credit note nets a sale only when
  customer, product, quantity and price all agree; partial returns and
  returns of pre-window sales find no match and simply drop out. Revenue is
  therefore net of clean cancellations but still slightly *gross* of messy
  returns.
- **A quarter of lines have no customer ID**
  (`r percent(mean(is.na(lines$customer_id)), accuracy = 0.1)` of clean lines
  — guest checkouts). They stay in revenue figures but cannot be used for
  customer-level analysis, which is therefore based on identified revenue
  only.

## Revenue over time (SQL)

Monthly totals come straight from SQLite — the cleaned lines are loaded into a
database and the business questions are answered in SQL
([`sql/queries.sql`](sql/queries.sql)). This dataset would fit in a data frame
comfortably; the point of the SQL layer is that these are exactly the queries
that live in a warehouse in production, and writing them as SQL keeps them
portable to one.

```{r monthly-table}
knitr::kable(
  sql$monthly_summary |>
    left_join(sql$monthly_growth |> select(month, mom_growth_pct), by = "month") |>
    mutate(
      revenue = label_gbp(revenue),
      # The first month has nothing to grow from; a bare NA in a rendered
      # table reads as an error, not an inapplicable cell.
      mom_growth_pct = coalesce(as.character(mom_growth_pct), "—")
    ),
  col.names = c("Month", "Revenue", "Invoices", "Identified customers",
                "Revenue / invoice (£)", "MoM growth (%)"),
  format.args = list(big.mark = ",")
)
```

```{r monthly-revenue, fig.height=4.0}
monthly <- sql$monthly_summary |>
  mutate(
    month_date = as_date(paste0(month, "-01")),
    # December 2011 stops on the 9th - nine trading days dressed up as a
    # month would read as a collapse, so the partial month is flagged.
    partial = month_date + months(1) - days(1) > span[2]
  )

ggplot(monthly, aes(month_date, revenue)) +
  geom_col(aes(alpha = partial), fill = PAL$blue, width = 22, show.legend = FALSE) +
  scale_alpha_manual(values = c(`FALSE` = 1, `TRUE` = 0.35)) +
  geom_text(
    data = filter(monthly, revenue == max(revenue)),
    aes(label = label_gbp(revenue)),
    vjust = -0.6, size = 3.2, colour = PAL$ink2
  ) +
  scale_y_continuous(labels = label_gbp, expand = expansion(mult = c(0, 0.12))) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b\n%Y") +
  labs(
    title = "Quiet for half the year, then a climb to the November peak",
    subtitle = "Monthly revenue, cleaned sales lines. The autumn ramp starts in September; November is the year's peak.",
    x = NULL, y = "Revenue",
    caption = "Faded bar: December 2011 is a partial month (data ends 9 December)."
  ) +
  theme_retail()
```

The shape is a giftware wholesaler's year in one picture: an unremarkable
first half, an autumn ramp as retail customers stock up for Christmas, a
`r sql$monthly_growth |> filter(month == "2011-11") |> pull(mom_growth_pct)`%
November jump — and a December bar that only looks like a crash because the
data stops on the 9th (its `trading_days` column in
[`output/monthly_growth.csv`](output/monthly_growth.csv) says the same to
anyone reading the CSV without this caption).

## The trading week

```{r weekday-profile, fig.height=3.6}
wd <- weekday_profile(lines)

ggplot(wd, aes(weekday, mean_daily_revenue)) +
  geom_col(fill = PAL$blue, width = 0.62) +
  scale_y_continuous(labels = label_gbp, expand = expansion(mult = c(0, 0.08))) +
  labs(
    title = "A weekday business: revenue peaks midweek, Saturdays are dark",
    subtitle = "Mean revenue per trading day, by weekday.",
    x = NULL, y = "Mean daily revenue",
    caption = "No Saturday trading appears anywhere in the data - the bar is absent, not zero."
  ) +
  theme_retail()
```

The peak day (`r as.character(wd$weekday[which.max(wd$mean_daily_revenue)])`)
averages `r label_gbp(max(wd$mean_daily_revenue))` in revenue, against
`r label_gbp(min(wd$mean_daily_revenue))` on the quietest day — consistent
with trade (B2B) customers ordering in working hours rather than consumers
browsing on weekends.

```{r daily-rolling, fig.height=4.0}
daily <- daily_revenue(lines)

ggplot(daily, aes(day, revenue)) +
  geom_point(colour = "#c3c2b7", size = 0.7) +
  geom_line(aes(y = rolling_7d), colour = PAL$blue, linewidth = 0.8, na.rm = TRUE) +
  scale_y_continuous(labels = label_gbp) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b\n%Y") +
  labs(
    title = "Daily revenue is noisy; the trend is not",
    subtitle = "Grey points: daily revenue. Blue line: 7-trading-day centred rolling mean.",
    x = NULL, y = "Daily revenue"
  ) +
  theme_retail()
```

Single-day spikes (a handful of very large wholesale orders) make the daily
series jumpy; the rolling mean shows the underlying level is flat until
August 2011 and then climbs for four straight months.

## What sells, and where (SQL)

```{r top-products-table}
knitr::kable(
  sql$top_products |> mutate(revenue = label_gbp(revenue)),
  col.names = c("Stock code", "Description", "Revenue", "Units", "Invoices"),
  format.args = list(big.mark = ",")
)
```

```{r top-products, fig.height=4.4}
top <- sql$top_products |>
  mutate(label = stringr::str_trunc(stringr::str_to_sentence(description), 34))

ggplot(top, aes(revenue, reorder(label, revenue))) +
  geom_col(fill = PAL$blue, width = 0.62) +
  scale_x_continuous(labels = label_gbp, expand = expansion(mult = c(0, 0.06))) +
  labs(
    title = paste0("Ten products out of ", comma(n_distinct(toupper(lines$stock_code))),
                   " earn ", percent(sum(sql$top_products$revenue) / total_revenue, accuracy = 0.1),
                   " of all revenue"),
    subtitle = "Top 10 stock codes by revenue, grouped case-insensitively.",
    x = "Revenue", y = NULL
  ) +
  theme_retail()
```

```{r countries, fig.height=3.8}
uk_share <- sql$country_summary |>
  filter(country == "United Kingdom") |> pull(revenue_share_pct)

intl <- sql$country_summary |>
  filter(country != "United Kingdom") |>
  slice_max(revenue, n = 8)

ggplot(intl, aes(revenue, reorder(country, revenue))) +
  geom_col(fill = PAL$blue, width = 0.62) +
  scale_x_continuous(labels = label_gbp, expand = expansion(mult = c(0, 0.06))) +
  labs(
    title = "Outside the UK, the near neighbours dominate",
    subtitle = paste0("Top international markets by revenue. The UK itself accounts for ",
                      uk_share, "% of all revenue."),
    x = "Revenue", y = NULL
  ) +
  theme_retail()
```

This is overwhelmingly a domestic business (`r uk_share`% UK), with the
Netherlands, Ireland, Germany and France making up most of the rest — the
kind of context that matters before anyone proposes "growing international
sales" off the back of one chart.

## Customer concentration

A *repeat* buyer here is a customer who bought on more than one calendar day
— a same-day flurry of invoices still counts as one visit. Guest checkouts
cannot be attributed to anyone, so this section covers identified revenue
only.

```{r repeat-table}
knitr::kable(
  sql$customer_repeat |> mutate(revenue = label_gbp(revenue)),
  col.names = c("Customer type", "Identified customers", "Revenue",
                "Share of identified revenue (%)"),
  format.args = list(big.mark = ",")
)
```

```{r pareto, fig.height=4.0}
pareto <- customer_pareto(lines)
top10_share <- pareto |> filter(customer_pct >= 0.10) |> slice(1) |> pull(revenue_share)

ggplot(pareto, aes(customer_pct, revenue_share)) +
  geom_line(colour = PAL$blue, linewidth = 0.9) +
  geom_vline(xintercept = 0.10, colour = PAL$muted, linetype = "dashed", linewidth = 0.4) +
  annotate(
    "text", x = 0.115, y = 0.30, hjust = 0, size = 3.2, colour = PAL$ink2,
    label = paste0("Top 10% of customers\n= ", percent(top10_share, accuracy = 1), " of identified revenue")
  ) +
  scale_x_continuous(labels = percent) +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  labs(
    title = "Revenue leans hard on a small core of customers",
    subtitle = "Cumulative share of identified revenue by customer rank (highest-spending first).",
    x = "Customers (ranked by spend)", y = "Cumulative revenue share",
    caption = "Identified customers only - guest-checkout revenue cannot be attributed."
  ) +
  theme_retail()
```

Repeat buyers are `r percent(sql$customer_repeat$customers[sql$customer_repeat$customer_type == "repeat"] / sum(sql$customer_repeat$customers), accuracy = 1)`
of identified customers but
`r sql$customer_repeat$revenue_share_pct[sql$customer_repeat$customer_type == "repeat"]`%
of identified revenue, and the top decile alone carries
`r percent(top10_share, accuracy = 1)`. Losing a handful of key accounts
would show up directly in the monthly numbers — concentration, not volume,
is this retailer's real commercial risk.

## A short-horizon forecast, with its caveats attached

```{r forecast, fig.height=4.2}
weekly <- weekly_revenue(lines)
fc <- forecast_weekly(weekly, h = 4)

ggplot() +
  geom_ribbon(
    data = fc$forecast, aes(week, ymin = lo95, ymax = hi95),
    fill = PAL$band, alpha = 0.55
  ) +
  geom_ribbon(
    data = fc$forecast, aes(week, ymin = lo80, ymax = hi80),
    fill = PAL$band
  ) +
  geom_line(data = fc$history, aes(week, revenue, colour = "All trading weeks"),
            linewidth = 0.5) +
  geom_line(data = fc$fit_run, aes(week, revenue, colour = "Model fit window"),
            linewidth = 0.8) +
  geom_line(
    data = fc$forecast, aes(week, mean, colour = "ETS forecast"),
    linewidth = 0.8, linetype = "21"
  ) +
  scale_colour_manual(
    NULL,
    values = c("All trading weeks" = "#c3c2b7", "Model fit window" = PAL$blue,
               "ETS forecast" = PAL$orange),
    breaks = c("All trading weeks", "Model fit window", "ETS forecast")
  ) +
  scale_y_continuous(labels = label_gbp) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b %Y") +
  labs(
    title = "Four weeks ahead: the level holds, the uncertainty is wide",
    subtitle = paste0("Weekly revenue with a 4-week ", fc$fit$method,
                      " forecast; bands are 80% / 95% intervals."),
    x = NULL, y = "Weekly revenue",
    caption = paste0(
      "Grey: every week the data window covers in full, including the year-end closure (near zero).\n",
      "The model is fit only on the unbroken run of weeks after that closure; ",
      "window-truncated edge weeks are excluded entirely."
    )
  ) +
  theme_retail() +
  theme(legend.position = c(0.17, 0.85), legend.background = element_blank())
```

Two modelling choices are doing the honest work here. The model is fit only
on the **unbroken run of weeks after the year-end closure**: `ts()` has no
notion of a calendar, so handing it the closure week (the grey plunge to
near zero) or the data window's truncated edge weeks would teach the level
a collapse that is really a holiday. And it is deliberately
**non-seasonal**: with thirteen months of history there is exactly one
observation of any annual pattern, so a seasonal term would be fit to
noise. That is also this forecast's honest limitation — it projects the
current level, and the *known* story of this business (the chart above it)
says December falls off a Christmas cliff that a one-year series cannot
teach a model. In production this would be the argument for pooling more
history before forecasting, not for a fancier model on the same data.

## Reproducing this document

```
Rscript R/get_data.R       # fetch + verify the raw data (sha256-pinned)
Rscript tests/run_tests.R  # unit tests for cleaning + SQL plumbing
Rscript run_analysis.R     # clean -> SQL -> tables -> knit this report
```

```{r session-info}
cat(sub("[[:blank:]]+$", "", capture.output(sessionInfo())), sep = "\n")
```
